The role of advection versus diffusion in intraventricular transport

### 2025-03-06 paper/main.tex
More updates in Methods

\subsection{Computational CSF dynamics}
The beating motion of the cilia generates CSF flow with steady rotational
structures~\cite{Olstad2019CiliaryDevelopment}
within the brain ventricles (\Crefpanels{fig:fig1}{a--c}).
In addition, cardiac pulsations induce pulsatile CSF flow.
We model this flow of CSF in the brain ventricles by the time-dependent,
incompressible Stokes equations which read as follows.
Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$
and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
      \rho \pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}
      \quad \text{in } \Omega,\\
      \nabla \cdot \uu &= 0
      \quad  \text{in } \Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$
and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. 
Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions.
Furthermore, $\rho=1000 \ \mathrm{kg/m^3}$ is the CSF density
and $\mu=0.70 \ \mathrm{mPa\cdot s}$ is the dynamic viscosity
of the CSF~\cite{Bloomfield1998EffectsFluid}.
We introduce the tangential traction $\bsigpar$ as the
tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$,
and $P_{\nn}(\rr)$ is the tangential projection of a vector $\rr$ onto $\Gamma$.
We consider three types of boundary conditions:
(i) \emph{tangential traction} (no flow normal to the boundary
$\uu \cdot \nn = 0$ with a stress $\btau$ applied tangentially on the boundary $\bsigpar = \btau$),
(ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in
the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$),
and (iii) \emph{free slip} (no flow normal to the boundary
$\uu \cdot \nn = 0$ and no tangential traction $\bsigpar = {\bf 0}$).
Unless noted otherwise, the free-slip condition is applied on the
boundary. We assume that the system starts at rest: $\uu(\xx, t=0) = \mathbf{0}$.

To study the impact of the motile cilia and the cardiac pulsations on the CSF flow,
we consider the following computational flow scenarios,
each predicting a CSF flow velocity $\uu$ and CSF pressure $p$.
\begin{itemize}
    \item A \emph{baseline} flow model representing a wild-type zebrafish including 
    contributions both from the cilia forces and the cardiac pulsations.
    In this scenario, we apply the tangential traction $\btau$ on the cilia boundary
    $\Gc$ and prescribe the normal pressure $\Tilde{p}$ on the anterior/posterior boundary $\Gp$. % Model 0/Model C

    \item A \emph{cilia-only/no-cardiac} flow model including contributions from the
    cilia forces but no cardiac pulsations.
    Here, we apply the tangential traction $\btau$ on the cilia boundary $\Gc$.
    In the absence of cardiac flow, the anterior/posterior boundary $\Gp=\emptyset$. % Model I/Model A

    \item A \emph{cardiac-only/no-cilia} flow model including contributions from the
    cardiac pulsations but no cilia forces.
    Here, we set the tangential traction to zero on the cilia boundary
    $\Gc$, while prescribing the normal pressure $\Tilde{p}$ on the
    anterior/posterior boundary $\Gp$.  % Model II/Model B
\end{itemize}

\subsubsection{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$ (\Crefpanels{fig:fig1}{d}, magenta markers),
we impose a constant-in-time tangential traction $\bsigpar=\btau$ to represent the net forces
of the cilia acting on the CSF.
We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$
and the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the
cilia region (see~\Cref{subsec:appendixA1}).
Both $\rr$ and the value $\tau=0.65$ mPa were chosen by 
model calibration using experimental data~\cite{Olstad2019CiliaryDevelopment}.
The data indicate speeds proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and
$2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the middle ventricle, respectively.
In the anterior ventricle, the speeds were estimated at $4.7 \pm 1.3$ \textmu m/s.
We calibrated $\tau$ heuristically by comparing these experimental velocity values
with the simulated, maximum velocity magnitude.

\subsubsection{Flow induced by cardiac pulsations}
To generate a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Crefpanels{fig:fig1}{d}, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s.
The amplitude $A=1.5$ mPa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
Similar to $\tau$, the amplitude $A$ was chosen by comparing experimental and simulated velocities.
Particle tracking data of pulsatile motion was used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (cf.~\Cref{sec:appC_particle_transport}).

>>>>

\subsection{Computational CSF dynamics in the brain ventricles}
The beating motion of the cilia generates CSF flow with persistent rotational
structures~\cite{Olstad2019CiliaryDevelopment}
within the brain ventricles (\Crefpanels{fig:fig1}{a--c}).
In addition, cardiac pulsations induce pulsatile CSF flow.
We model this flow of CSF by the time-dependent,
incompressible Stokes equations which read as follows.
Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$
and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
      \rho \pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}
      \quad \text{in } \Omega,\\
      \nabla \cdot \uu &= 0
      \quad  \text{in } \Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$
and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. 
Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions.
Furthermore, $\rho=1000 \ \mathrm{kg/m^3}$ is the CSF density
and $\mu=0.70 \ \mathrm{mPa\cdot s}$ is the dynamic viscosity
of the CSF~\cite{Bloomfield1998EffectsFluid}.
We introduce the tangential traction $\bsigpar$ as the
tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$,
and $P_{\nn}(\rr)$ is the tangential projection of a vector $\rr$ onto $\Gamma$.
We consider three types of boundary conditions:
(i) \emph{tangential traction} (no flow normal to the boundary
$\uu \cdot \nn = 0$ with a stress $\btau$ applied tangentially on the boundary $\bsigpar = \btau$),
(ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in
the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$),
and (iii) \emph{free slip} (no flow normal to the boundary
$\uu \cdot \nn = 0$ and no tangential traction $\bsigpar = {\bf 0}$).
Unless noted otherwise, the free-slip condition is applied on the
boundary. We assume that the system starts at rest: $\uu(\xx, t=0) = \mathbf{0}$.

To study the impact of the motile cilia and the cardiac pulsations on the CSF flow,
we consider the following computational flow scenarios,
each predicting a CSF flow velocity $\uu$ and pressure $p$.
\begin{itemize}
    \item A \emph{baseline} flow model representing a wild-type zebrafish including 
    contributions both from the cilia forces and the cardiac pulsations.
    In this scenario, we apply the tangential traction $\btau$ on the cilia boundary
    $\Gc$ and prescribe the normal pressure $\Tilde{p}$ on the anterior/posterior boundary $\Gp$. % Model 0/Model C

    \item A \emph{cilia-driven/no-cardiac} flow model including contributions from the
    cilia forces but no cardiac pulsations.
    Here, we apply a tangential traction $\btau$ on the cilia boundary $\Gc$.
    In the absence of cardiac flow, the anterior/posterior boundary $\Gp=\emptyset$. % Model I/Model A

    \item A \emph{cardiac-induced/no-cilia} flow model including contributions from the
    cardiac pulsations but no cilia forces.
    Here, we set the tangential traction to zero on the cilia boundary
    $\Gc$, while prescribing the normal pressure $\Tilde{p}$ on the
    anterior/posterior boundary $\Gp$.  % Model II/Model B
\end{itemize}
The tangential traction and normal pressure are described further below.

\subsubsection{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$
(\Crefpanels{fig:fig1}{d}, magenta markers), we impose a
constant-in-time tangential traction $\bsigpar=\btau$ to represent the
net forces of the cilia acting on the CSF.  We set $\btau(\xx, \rr) =
\tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$ and
the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the
cilia region (see~\Cref{subsec:appendixA1}). Previous
experiments~\cite{Olstad2019CiliaryDevelopment} indicate flow speeds
proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and
$2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the
middle ventricle, respectively.  In the anterior ventricle, the speeds
were estimated at $4.7 \pm 1.3$ \textmu m/s. We calibrated $\rr$ and
$\tau=0.65$ mPa by comparison of the simulated maximum velocity
magnitude with these flow speeds.

%% comparing these experimental velocity values
%% with the simulated, maximum velocity magnitude.

%% Both $\rr$ and the value $\tau=0.65$ mPa were chosen by 
%% calibration against experimentally measured flow speeds~\cite{Olstad2019CiliaryDevelopment}.

\subsubsection{Flow induced by cardiac pulsations}
To generate a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Crefpanels{fig:fig1}{d}, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
we set the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s. The amplitude $A=1.5$ mPa determines the magnitude of the pulsatile flow
and was calibrated based on previous experiments~\cite{Olstad2019CiliaryDevelopment}.
Particle tracking data were used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (cf.~\Cref{sec:appC_particle_transport}).
